\documentclass{article}
\usepackage{geometry}
\geometry{a4paper, margin=1in}
\usepackage{graphicx}
\usepackage[colorlinks=true, linkcolor=blue, citecolor=blue, urlcolor=blue]{hyperref}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{amsmath}
\usepackage{enumitem}

\lstset{
    language=Python,
    basicstyle=\ttfamily\footnotesize\selectfont, % use the selected monospaced font
    backgroundcolor=\color{white},
    keywordstyle=\color{blue},
    commentstyle=\color{gray},
    stringstyle=\color{red},
    numbers=left,
    numberstyle=\tiny\color{gray},
    stepnumber=1,
    numbersep=10pt,
    frame=single,
    breaklines=true,
    captionpos=b,
    tabsize=4
}

\title{Progress Report - Week 3 \\
\large Novel Methods for Determining Neuron Sensitivity in Large Language Models}
\author{
    [Welby Seely] \\
    \texttt{[wseely@emich.edu]}
}
\date{\today}

\begin{document}

    \maketitle

    \section{Project Overview}\label{sec:project-overview}
    Large Language Model (LLM) power consumption is increasingly problematic for datacenters, with \textit{The Electric Power
    Research Institute} forecasting that data centers may see their electricity consumption double by
    2030~\cite{kindig2024}.

    Optimizing these models is key to reducing power consumption.
    The cornerstone of optimization is determining what you need, and what you don't: determine the importance of
    aspects and components of the system, enabling you to perform systematic optimization with certainty.
    In Artificial Neural Networks (ANN), the atomic unit of these components is the artificial neuron.
    To optimize an ANN, understanding the sensitivity of the neuron to changes in its input and parameters is paramount,
    but the sheer size and complexity of LLMs makes this incredibly challenging.

    Thus, the goal of this research is to:
    \begin{enumerate}
        \item Understand current methodologies for determining neuron sensitivity in LLMs.
        \item Explore improvements that can be made to more accurately and more efficiently determine neuron sensitivity in
        LLMs.
    \end{enumerate}

    The intent is that this research will be a first step towards innovating and improving optimizations for LLMs,
    reducing power consumption and enabling the models to run on lower powered hardware.

    \section{Current Progress}\label{sec:current-progress}
    \subsection{Comparative Analysis of Neuron Sensitivity Techniques}\label{subsec:comparative-analysis-of-neuron-sensitivity-techniques}
    \input{structure-analysis}
    \input{token-importance}
    \subsection{Neuron Sensitivity Techniques to Be Analyzed}\label{subsec:neuron-sensitivity-techniques-to-be-analyzed}
    \begin{itemize}
        \item Mechanistic Interpretability
        \subitem Maps out neural circuits in order to determine which parts of a model are responsible for specific capabilities (e.g.\ arithmetic, semantic relationships between words)~\cite{singh2024rethinking}.
        \item Activation Maximization
        \subitem Optimize an input to maximize the activation of a specific neuron or neuron layer~\cite{erhan2009visualizing}.
        \item Saliency Maps
        \subitem Visualize Neuron sensitivity by highlighting inputs that most affect a particular neuron or output~\cite{hsu2023explainable}.
        \item Layer-wise Relevance Propagation
        \subitem Similar to Saliency Maps, decomposes the prediction of a network back to the individual input features~\cite{jia2022interpreting}.
        \item Integrated Gradients
        \subitem Quantifies feature importance by integrating gradients along the path from a baseline input to the actual input~\cite{sundararajan2017axiomatic}.
        \item SHAP Values (SHapley Additive exPlanation)
        \subitem Quantifies sensitivity by assigning a contribution score to each input feature based on its impact on the model’s output~\cite{nohara2022explanation}.
    \end{itemize}

    \section{Challenges and Issues}\label{sec:challenges-and-issues}
    Everything is perfect, no worries here.
    I have truly mastered all things Machine Learning and am now ready to trascend nature and become one with artificial intelligence.

    \section{Next Steps}\label{sec:next-steps}
    \begin{enumerate}
        \item Continue with comparative analysis of methods.
        Will try not to get too bogged down on the specifics of each paper in order to get a better view of the bigger picture.
        \item Expand research on ``Coupled Structure Analysis'' which holds particular promise.
        \item Begin familiarization with TensorBoard, in particular with LLM models in the HuggingFace Transformers library.
    \end{enumerate}

    \section{Questions or Feedback Needed}\label{sec:questions-or-feedback-needed}
    Based on your feedback, I'll be completing an initial comparative analysis of sensitivity/importance methods in the next week or two.
    I think I need to then narrow the scope of the research in order to have a ``minimum viable product'' by the end of the semester.
    Do you have any advice on how to best spend my time?
    I suspect I'm getting too bogged down on specific papers as I read them.

    \bibliographystyle{plainurl}
    \bibliography{bibliography}
    \thanks Thanks to ChatGPT~\cite{chatgpt_2024} for helping me understand papers as I parse through them and generating
    BibTeX citations for websites.

\end{document}
