% Feedback:
% Start actually implementing (or even better try to find open-source implementations) models
% How do these works evaluate their energy savings (if they do)? Do they evaluate actual hardware/simulated hardware?

\subsubsection{Structure Importance Analysis}\label{subsubsec:strucure-importance-analysis}

This is an analysis of optimization techniques that rely on understanding on the importance of larger structures with the LLM\@.
\begin{itemize}
    \item Llm-pruner: On the structural pruning of large language models~\cite{ma2023llm}.
    \begin{itemize}
        \item Identifies dependency-based structures, in the context of pruning
        \item Axiom: a neuron $N_j$ is co-dependent on another neuron $N_i$ if and only if it only receives input from $N_j$ and $N_i$ only outputs to $N_j$.
        More formally:
        \begin{equation}
            N_j \in \operatorname{Out}(N_i) \wedge \operatorname{Deg}^-(N_j) = 1 \Rightarrow N_j \text{ is dependent on } N_i\label{eq:equation}
        \end{equation}
        \begin{equation}
            N_i \in \operatorname{In}(N_j) \wedge \operatorname{Deg}^+(N_i) = 1 \Rightarrow N_i \text{ is dependent on } N_j\label{eq:equation2}
        \end{equation}
        \item Identifying groups (building a dependency graph):
        \begin{enumerate}
            \item Select an arbitrary neuron as an initial ``trigger.''
            \item Using the axiom, identify if there is a dependency relationship with another neuron, either forwards or backwards.
            \item Newly identified neurons trigger additional ``dependent'' neurons, if eligible by the axiom.
            \item Continue this process iteratively until no new dependent neurons are found.
        \end{enumerate}
        \item Estimating the importance of a group:
        \begin{enumerate}
            \item Identify a small, publicly available dataset (different from the training data).
            \item Measure the importance of a group by using Vector-Wise and Element Importance in aggregate:
            \begin{enumerate}
                \item Vector-Wise Importance: relate weight matrices to the loss function.
                This is useful because using this non-training data means ${\partial \mathcal{L}}/{\partial W_i} \not \approx 0$.
                \item Element Wise Importance: relate individual parameters to the loss function.
                \item Finally, these two importance scores must be aggregated through a summation, a maximum, a product, or the last parameter in the group.
            \end{enumerate}
        \end{enumerate}
        \item Advantages:
        \begin{itemize}
            \item The simplistic nature of the axiom makes it easy to identify groups eligible for simple pruning.
            \item Element-wise importance calculation is simplified through the use of approximate gradients and Hessians.
            \item The small dataset keeps the calculations to a relative minimum.
        \end{itemize}
        \item These advantages would result in reduced power consumption, but the paper unfortunately does not delve into these details.
        More research into pruning methods is necessary.
        \item Weaknesses:
        \begin{itemize}
            \item The grouping algorithm cannot classify all neurons into groups - ``dependence'' is relegated to a very simplistic definition that prevents us from looking at the entire model from a holistic point of view.
            \item The paper reports a rapid collapse in model accuracy and coherence at $~20\%$ pruning of neurons.
            I suspect that this is a limitation of the axiom, the naive definition of ``dependence'':
            the actual importance of groups of neurons cannot truly be understood when removing the more complex neural relationships from the analysis.
            This is a possible narrowing of focus this research could take.
            \item The grouping algorithm is not foolproof, and fails for certain models and ``operations''\cite{LLM-Pruner}.
        \end{itemize}
        \item Energy usage and savings
        \begin{itemize}
            \item Energy savings not directly evaluated by the authors.
            Tests likely carried out using enterprise grade GPUs.
        \end{itemize}
    \end{itemize}
    \item LLM Pruning and Distillation in Practice: The Minitron Approach~\cite{sreenivas2024llm}
    \begin{itemize}
        \item Uses activation-based importance estimation strategy that computes sensitivity information for all
        the axes under consideration (depth, neuron, head, and embedding channel).
        \item Energy usage and savings
        \begin{itemize}
            \item Energy savings not evaluated by the authors.
            \item Measuring on local hardware: energy consumption when compared to baseline Llama 3 model are
            approximately 23\% per token (3.17 vs 4.12 J).
        \end{itemize}
    \end{itemize}
    \item Rethinking Interpretability in the Era of Large Language Models ~\cite{singh2024rethinking}
    \begin{itemize}
        \item Paper is something of a meta-analysis of mechanistic interpretability.
        The fundamental idea is to map out neural circuits in order to determine which parts of a model are responsible for specific capabilities (e.g.\ arithmetic, semantic relationships between words).
        Over 150 papers cited in this meta-analysis.
        \item Energy usage and savings
        \begin{itemize}
            \item Energy savings not evaluated by the authors.
        \end{itemize}
    \end{itemize}
\end{itemize}